---
title: "Age-at-risk expansion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Age-at-risk expansion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  class.output = "outp",
  class.message = "outp",
  class.warning = "outp",
  class.error = "outp",
  comment = "#"
)
library(ckbtools)
```

# Introduction

Analysis of time-to-event data (survival analysis) is often used estimate associations between exposures and disease outcomes. The Cox proportional hazards model is widely used in such analyses. This model has the key "proportional hazards" assumption that effect of covariates (hazard ratios) are constant over time.

Age is frequently a confounder of the association between exposure and disease, or a covariate of interest itself. When time-on-study is used as the time scale for a Cox model, age at baseline can be adjusted for but this assumes a constant effect of baseline age over time and does not account for the effect of 'acquired' age during follow-up.

Lexis expansion involves splitting each participant's follow-up time into intervals. Lexis expansion by age-at-risk will create observations for each subject for each age band (e.g. 5-year intervals) they pass through during follow-up. For example, using 5-year age bands between 30 years and 80 years, a participant who enters the study at age 42 years and leaves age 53 years will have three rows of data: from baseline until their 45th birthday; from their 45th birthday to 50th birthday; and from their 50th birthday to the end of follow-up.

The newly created age-at-risk bands can be used as a stratification variable. When the effect of age-at-risk is not the primary focus, baseline hazard can be stratified by age-at-risk. Each age-at-risk band then has its own baseline hazard. The effect of all other covariates is assumed to be constant across age-at-risk bands and throughout follow-up.

It may of interest to estimate an effect for a covariate in each age-at-risk band. This can be achieved by including an interaction between age-at-risk band and the covariate. Further, an additional nuisance term nested within age-at-risk band can be included to avoid assuming that proportional associations are similar across different, wider, bands of age. See [Multi-age group](#multi-age-group).

The following examples demonstrates how to implement such models in R, using the `expand_age_at_risk()` function to apply Lexis expansion for age-at-risk.


# Examples

## Expand by age-at-risk

The data frame contains one row per participant. For example:

```{r}
ckbtools_participant_data[ckbtools_participant_data$csid == "1000005",
                          c("csid", "study_date", "dob_anon", "ihd", "ihd_date")]
```

To expand our data frame we specify the age-at-risk-bands, and identify the columns that contain participants' unique ID (default: "csid"), date of birth ("dob_anon"), date of study entry ("study_date"), date of study exit ("endpoint_date") and status at exit ("endpoint").

The age-at-risk strata are chosen by specifying a vector of cut points. The first will be the minimum age-at-risk to enter the analysis, and the last will be the age at which participants leave the analysis. For example, `seq(from = 40, to = 80, by = 5)` will use 5-year age-at-risk strata bands from 40 to 80 years.

```{r}
# Expand data frame by age-at-risk
participant_data_long <- expand_age_at_risk(data          = ckbtools_participant_data,
                                            col.exit_date = "ihd_date",
                                            col.status    = "ihd",
                                            ages          = seq(from = 40, to = 80, by = 5),
                                            col.keep      = "bmi_grp")
```

The expanded data contains multiple rows per participant: one for each age-at-risk strata during which they have some follow-up time. For example:

```{r}
participant_data_long[participant_data_long$csid == "1000005",
                      c("csid", "agegrp",
                        "start_int", "end_int",
                        "status", "t_start_days", "t_end_days")]
```

The expanded data frame contains the participant ID columns used to expand the data (set by arguments `col.id`, `col.dob`, `col.entry_date`, `col.exit_date`, `col.status`) and new columns:

- **agegrp_start**: Lower bound for age-at-risk strata
- **agegrp_end**: Upper bound for age-at-risk strata
- **agegrp**: Age-at-risk strata label
- **start_int**: Start date for time interval
- **end_int**: End date for time interval
- **status**: Participant status at end of age-at-risk interval
- **t_start_days**: Start time for interval (days)
- **t_end_days**: End time for interval (days)
- **time_in**: Start time for interval (years, calculated as days/365.25)
- **time_out**: End time for interval (years, calculated as days/365.25)


## Cox proportional hazards model stratified by age-at-risk

The expanded data frame contains column `agegrp` which can be used to stratify the Cox proportional hazards model by including a `strata(agegrp)` in the model formula and using the time columns to specify the starting and ending time for intervals.

```{r}
library(survival)
fit <- coxph(Surv(time_in, time_out, status) ~ bmi_grp + strata(agegrp),
             data = participant_data_long)
broom::tidy(fit)
```

```{r}
print(fit$xlevels$`strata(agegrp)`)
```

## Multi-age group

The same expanded dataset can be used for other models, such as those estimating HRs with an interaction between the risk factor and age-at-risk. For example, an interaction with 10-year age-at-risk band can be included while allowing hazards to differ between the first and second half of that band. By including an additional nuisance term nested within age band, the HR in each 10-year band is estimated as the geometric mean of the HRs in the first and second half of that decade, avoiding assumptions that proportional associations are similar across different decades of age. See [Lacey et al., 2018](https://doi.org/10.1016/S2214-109X(18)30217-1) for an example with major vascular disease and systolic blood pressure.

To fit such a model, first create a new column from `agegrp_start` called `age_band` that is the 10-year age-at-risk band. Then create a numeric column `age_band_midpoint` that is the midpoint of each decade. Also create a column `agegrp_midpoint` that is the mid-point of each 5-year age-at-risk group. The nuisance term `nuis` is the different between the midpoint of the 10-year age-at-risk band and the 5-year age-at-risk group. Finally, create a column that is the interaction between risk factor group and 10-year age-at-risk band (here, `bmi_grp_x_age_band` for BMI group).

```{r}
participant_data_long <- participant_data_long %>%
  dplyr::mutate(age_band = cut(agegrp_start,
                               breaks = c(40, 50, 60, 70, 80),
                               labels = c("40-49", "50-59", "60-69", "70-79"),
                               right = FALSE)) %>%
  dplyr::mutate(age_band_midpoint = dplyr::case_match(age_band,
                                                      "40-49" ~ 45,
                                                      "50-59" ~ 55,
                                                      "60-69" ~ 65,
                                                      "70-79" ~ 75),
                agegrp_midpoint = (agegrp_start + agegrp_end) / 2,
                nuis = age_band_midpoint - agegrp_midpoint,
                bmi_grp_x_age_band = interaction(bmi_grp, age_band, sep = "_"))
```

In the Cox model formula, include the BMI group by 10-year age-at-risk band interaction (`bmi_grp_x_age_band`) and the nuisance term nested in 10-year age-at-risk bands (`nuis:age_band`).

```{r}
fit <- coxph(Surv(time_in, time_out, status) ~ bmi_grp_x_age_band + nuis:age_band,
             data = participant_data_long)
```

A log HR is estimated for each combination of BMI group and 10-year age-at-risk band. The reference group is BMI \<18.5 and age-at-risk 40-49, and so this group has a hazard ratio of one.

```{r}
broom::tidy(fit)
```

# Note

The end times for each age-at-risk strata (columns t_end_days and time_out) are always x.95. This helps avoid ambiguity about risk sets, but the time between the start and end times does not equal a whole number of days. So these columns should not be used to calculate rates or time at risk.


# Using cox_model() and multiple_models()

The Cox model can also be fit using the `cox_model()` function in this package with the expanded data frame

```{r}
cox_model(
  participant_data_long,
  exposure   = "bmi_grp",
  outcome    = "status",
  time_in    = "time_in",
  time_out   = "time_out",
  adjust_for = "strata(agegrp)",
  return     = "coefs"
)
```

If using the `multiple_models()` function, age-at-risk expansion and stratification can be used via the `age_at_risk` argument. A list of arguments for the `expand_age_at_risk()` function is supplied as a named list.

Note the it is the original non-expanded data set that must be used, and that `"strata(agegrp)"` is automatically added to the model formula.

```{r}
results <- multiple_models(
  ckbtools_participant_data,
  model       = "cox",
  exposures   = "bmi_grp",
  outcomes    = "ihd",
  age_at_risk = list(ages = seq(from = 40, to = 80, by = 5)),
  return      = "coefs",
  verbose     = "quiet"
)

results[, c("term", "estimate", "std.error", "n_group", "nevent_group")]
```

If there is only one outcome, then `multiple_models()` can be used with the expanded data frame, but the arguments `outcomes`, `time_in` and `time_out` must be set appropriately.

```{r}
results <- multiple_models(
  participant_data_long,
  model      = "cox",
  exposures  = "bmi_grp",
  outcomes   = "status",
  time_in    = "time_in",
  time_out   = "time_out",
  adjust_for = list(none = "strata(agegrp)"),
  return     = "coefs",
  verbose     = "quiet"
)

results[, c("term", "estimate", "std.error", "n_group", "nevent_group")]
```


# References

[San H, Lewington S. Lexis expansion-age-at-risk adjustment for survival analysis. InBrussels: Statistics and Pharmacokinetics Pharmaceutical Users Software Exchange conference 2013 Oct 13.](https://www.lexjansen.com/phuse/2013/sp/SP09.pdf)

[Moolgavkar, S. H., Chang, E. T., Watson, H. N. & Lau, E. C. An Assessment of the Cox Proportional Hazards Regression Model for Epidemiologic Studies. Risk Analysis 38, 777–794 (2018).](https://doi.org/10.1111/risa.12865)

[Vyas, M. V., Fang, J., Kapral, M. K. & Austin, P. C. Choice of time-scale in time-to-event analysis: evaluating age-dependent associations. Annals of Epidemiology 62, 69–76 (2021).](https://doi.org/10.1016/j.annepidem.2021.06.006) (Not open access)

[Hurley, M. A. A reference relative time-scale as an alternative to chronological age for cohorts with long follow-up. Emerg Themes Epidemiol 12, 18 (2015).](https://doi.org/10.1186/s12982-015-0043-6)

[Canchola AJ, Stewart SL, Bernstein L, West DW, Ross RK, Deapen D, Pinder R, Reynolds P, Wright W, Anton-Culver H, Peel D. Cox regression using different time-scales. Western Users of SAS Software. San Francisco, California. 2003 Nov 5.](https://www.lexjansen.com/wuss/2003/DataAnalysis/i-cox_time_scales.pdf)

[Lacey, B. et al. Age-specific association between blood pressure and vascular and non-vascular chronic diseases in 0·5 million adults in China: a prospective cohort study. The Lancet Global Health 6, e641–e649 (2018).](https://doi.org/10.1016/S2214-109X(18)30217-1)
