---
title: "Associations of proteins in a case-subcohort study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Associations of proteins in a case-subcohort study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  class.output = "outp",
  class.message = "outp",
  class.warning = "outp",
  class.error = "outp",
  comment = "#"
)
library(ckbtools)

# set number of Olink plates in synthetic data to prevent warning about number
# of plates
options(olink_expected_plates = 26)
```


This article demonstrates how to use this package to estimate associations between proteins and risk of incidence disease using Cox proportional hazards models, where the study uses a case-subcohort design. For a real application, see ["Plasma Proteomics to Identify Drug Targets for Ischemic Heart Disease"](https://doi.org/10.1016/j.jacc.2023.09.804).


# Import data

First, all relevant participant and protein data must be imported into R.
For example, use `readRDS()` for RDS format data or `read.csv()` for csv data.
```{r, eval = FALSE}
baseline       <- readRDS("baseline.rds")
ascertainments <- readRDS("ascertainments.rds")
endpoints      <- readRDS("endpoints.rds")

raw_olink_meta_data <- readRDS("olink_explore_meta.rds")
raw_olink_protein_data <- readRDS("data_baseline_olink_explore.rds")

raw_somascan_meta_data <- readRDS("somalogic_meta.rds")
raw_somascan_protein_data <- readRDS("data_baseline_somalogic.rds")
raw_somascan_sample_data  <- readRDS("data_baseline_somalogic_samples.rds")
```

For demonstration, the synthetic data sets available in this package can be used:

1. ckbtools_participant_data
2. ckbtools_ascertainments
3. ckbtools_raw_olink_meta_data
4. ckbtools_raw_olink_protein_data
5. ckbtools_raw_somascan_meta_data
6. ckbtools_raw_somascan_protein_data
7. ckbtools_raw_somascan_sample_data

## Baseline and endpoint data

A data set containing relevant participant baseline and endpoint data must be prepared manually. See [CKB data](ckb_data.html) for an example.

In this example, the outcome is IHD.

## Identify case-subcohort participants

```{r}
ckbtools_participant_data$subcohort <- ckbtools_participant_data$proteomics_subcohort

## keep only the subcohort and IHD cases outside the subcohort
ckbtools_participant_data <- ckbtools_participant_data[ckbtools_participant_data$subcohort == 1 |
                                                         ckbtools_participant_data$ihd == 1, ]
```

## Prentice weighting

To use the [Prentice weighting method for case-subcohort designs](https://doi.org/10.1093/biomet/73.1.1), follow up time starts at zero for participants in the subcohort but starts just before outcome for cases outside subcohort.

```{r}
ckbtools_participant_data$time_out <- as.numeric(ckbtools_participant_data$ihd_date - ckbtools_participant_data$study_date)
ckbtools_participant_data$time_in <- ckbtools_participant_data$time_out - 0.001
ckbtools_participant_data$time_in[ckbtools_participant_data$subcohort == 1] <- 0
```

# Prepare protein data

Prepare the protein data as described in [Proteomics](proteomics.html#preparing-protein-data).

For this analysis, no plate correction is applied to the protein values but we will adjust for "plateid" in the analysis models. SomaScan values are log<sub>2</sub> transformed. Protein values are then scaled to have standard deviations of one.

```{r, message=FALSE}
## set CKB data release
options(ckb.data.release = "19.02")

## meta data
olink_meta_data <- prepare_olink_meta_data(ckbtools_raw_olink_meta_data,
                                           ckbtools_raw_olink_protein_data)

somascan_meta_data <- prepare_somascan_meta_data(ckbtools_raw_somascan_meta_data)

## protein data
olink_protein_data <- prepare_olink_data(ckbtools_raw_olink_protein_data,
                                         olink_meta_data)

somascan_protein_data <- prepare_somascan_data(ckbtools_raw_somascan_protein_data,
                                               ckbtools_raw_somascan_sample_data,
                                               somascan_meta_data,
                                               transform = "log2")

## proteomics analysis data
olink_analysis_data <- prepare_proteomics_analysis_dataset(
  ckbtools_participant_data,
  olink_protein_data,
  olink_meta_data,
  plate_correction = "none",
  scaling = "scale"
)

somascan_analysis_data <- prepare_proteomics_analysis_dataset(
  ckbtools_participant_data,
  somascan_protein_data,
  somascan_meta_data,
  plate_correction = "none",
  scaling = "scale"
)
```


# Analysis 

The `multiple_models()` function can be used to run multiple regression models.

*It is highly recommended to first run some individual analyses  to check that the models and output are as intended, before running analyses for many proteins.*

The code below fits models for all Olink and SomaScan proteins. To set the correct entry and exit times for the Prentice weighting use the `time_in` and `time_out` arguments. Robust standard errors should also be used by setting `clusted = "csid"`.

```{r, message=FALSE}
fully_adjusted <- c("strata(sex)",
                    "strata(region)",
                    "poly(age_at_study_date, 2)",
                    "poly(hours_since_last_ate, 2)",
                    "poly(region_mean_temp, 2)",
                    "education",
                    "smoking",
                    "alcohol",
                    "met",
                    "sbp_mean",
                    "has_diabetes",
                    "bmi_calc",
                    "plateid")


olink_results <- multiple_models(
  olink_analysis_data,
  model      = "cox",
  exposures  = olink_meta_data$name,
  outcomes   = "ihd",
  time_in    = "time_in",
  time_out   = "time_out",
  cluster    = "csid",
  adjust_for = list(full = fully_adjusted)
)

somascan_results <- multiple_models(
  somascan_analysis_data,
  model      = "cox",
  exposures  = somascan_meta_data$name,
  outcomes   = "ihd",
  time_in    = "time_in",
  time_out   = "time_out",
  cluster    = "csid",
  adjust_for = list(full = fully_adjusted)
)
```

*Note that education, smoking and alcohol are formatted as factors in the data, so they are correctly used as categorical variables in the model. If you want to include a categorical variable that is coded numerically then make sure to use `"as.factor(variable)"`.*

# Summarise results

## Check for errors and warnings

After fitting multiple models, always check for warnings and errors. If there are any, decide how these will be handled when summarising results or for further analyses.

```{r}
olink_results %>%
  dplyr::filter(!is.na(error) | !is.na(warnings)) %>%
  dplyr::distinct(outcome, exposure, adjustment, error, warnings)

somascan_results %>%
  dplyr::filter(!is.na(error) | !is.na(warnings)) %>%
  dplyr::distinct(outcome, exposure, adjustment, error, warnings)
```

## Select one result per protein

Some Olink assays are repeated across panels. To summarise the results as in the referenced paper, select the assay by panel preference Cardiometabolic > Inflammation > Neurology > Oncology, and prefer models with no error or warning.

```{r}
olink_summary_results <- olink_results %>%
  dplyr::filter(exposure %in% olink_meta_data$name) %>%
  dplyr::inner_join(olink_meta_data, by = c("exposure" = "name")) %>%
  dplyr::group_by(outcome, adjustment, uniprot) %>%
  dplyr::arrange(!is.na(error) | !is.na(warnings), panel) %>%
  dplyr::filter(dplyr::row_number() == 1) %>%
  dplyr::ungroup()
```

Some SomaScan aptamers target the same protein. To summarise the results as in the referenced paper, select one aptamer per target choosing the result with the smallest p-value and no error or warning (or just the smallest p-value if all have error or warning)

```{r}
somascan_summary_results <- somascan_results %>%
  dplyr::inner_join(somascan_meta_data, by = c("exposure" = "name")) %>%
  dplyr::group_by(adjustment, target) %>%
  dplyr::arrange(!is.na(error) | !is.na(warnings), p.value) %>%
  dplyr::filter(dplyr::row_number() == 1) %>%
  dplyr::ungroup()
```


## Multiple testing

In this example, apply a false discovery rate correction separately in the Olink and SomaScan analyses.

```{r}
olink_summary_results <- olink_summary_results %>%
  ## handle errors and warnings
  dplyr::mutate(p.value = dplyr::if_else(!is.na(error) | !is.na(warnings), 1, p.value)) %>%
  ## multiple testing correction
  dplyr::mutate(p.value.fdr = p.adjust(p.value, method = "fdr"))

somascan_summary_results <- somascan_summary_results %>%
  ## handle errors and warnings
  dplyr::mutate(p.value = dplyr::if_else(!is.na(error) | !is.na(warnings), 1, p.value)) %>%
  ## multiple testing correction
  dplyr::mutate(p.value.fdr = p.adjust(p.value, method = "fdr"))
```


# Present results

The number of proteins significantly associated (FDR-corrected p < 0.05) for Olink and SomaScan proteins can then be summarised.

```{r}
olink_summary_results %>%
  dplyr::summarise(n = dplyr::n(),
                   error_or_warning = sum(!is.na(error) | !is.na(warnings)),
                   significant = sum(p.value.fdr < 0.05),
                   significant_inverse = sum(p.value.fdr < 0.05 & estimate < 0),
                   .groups = "drop")

somascan_summary_results %>%
  dplyr::summarise(n = dplyr::n(),
                   error_or_warning = sum(!is.na(error) | !is.na(warnings)),
                   significant = sum(p.value.fdr < 0.05),
                   significant_inverse = sum(p.value.fdr < 0.05 & estimate < 0),
                   .groups = "drop")
```

```{r}
olink_summary_results %>%
  dplyr::arrange(p.value.fdr) %>%
  dplyr::select(assay, uniprot, estimate, robust.se, p.value, p.value.fdr) %>%
  head(5)

somascan_summary_results %>%
  dplyr::arrange(p.value.fdr) %>%
  dplyr::select(target, uniprot, estimate, robust.se, p.value, p.value.fdr) %>%
  head(5)
```
