---
title: "Associations of proteins with participant characteristics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Associations of proteins with participant characteristics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  class.output = "outp",
  class.message = "outp",
  class.warning = "outp",
  class.error = "outp",
  comment = "#"
)
library(ckbtools)

# set number of Olink plates in synthetic data to prevent warning about number
# of plates
options(olink_expected_plates = 26)
```

This article demonstrates how to use this package to estimate cross-sectional associations between proteins and participant characteristics. The methods used are based on the papers ["Associations of 2923 Olink proteins with demographic, lifestyle, environmental and health characteristics in middle-aged Chinese adults"](https://doi.org/10.1007/s10654-025-01311-z) and ["Associations of 6600 SomaScan proteins with demographic, lifestyle, environmental and health characteristics in Chinese adults"](https://doi.org/10.1038/s41598-026-41444-z).


# Import data
First, all relevant participant and protein data must be imported into R. For example, use `readRDS()` for RDS format data or `read.csv()` for csv data.

```{r, eval = FALSE}
baseline       <- readRDS("baseline.rds")
ascertainments <- readRDS("ascertainments.rds")

raw_olink_meta_data <- readRDS("olink_explore_meta.rds")
raw_olink_protein_data <- readRDS("data_baseline_olink_explore.rds")

raw_somascan_meta_data <- readRDS("somalogic_meta.rds")
raw_somascan_protein_data <- readRDS("data_baseline_somalogic.rds")
raw_somascan_sample_data  <- readRDS("data_baseline_somalogic_samples.rds")
```

For demonstration, the synthetic data sets available in this package can be used:

1. ckbtools_participant_data
2. ckbtools_ascertainments
3. ckbtools_raw_olink_meta_data
4. ckbtools_raw_olink_protein_data
5. ckbtools_raw_somascan_meta_data
6. ckbtools_raw_somascan_protein_data
7. ckbtools_raw_somascan_sample_data



# Preparation of participant data

A data set containing relevant participant baseline data must be prepared manually. For these analyses, the data must contain all participant characteristics used for  exclusions or as adjustment variables. See [CKB data](ckb_data.html) for an example.


## Select subcohort participants

These analyses of baseline characteristics will use only participants in the subcohort of the nested case-subcohort proteomics study.

```{r}
subcohort_participant_data <- ckbtools_participant_data[ckbtools_participant_data$proteomics_subcohort == 1, ]
subcohort_participant_data$subcohort <- 1
```


# Prepare protein data

Prepare the protein data as described in [Proteomics](proteomics.html#preparing-protein-data).

```{r}
options(ckb.data.release = "19.02")

## meta data
olink_meta_data <- prepare_olink_meta_data(ckbtools_raw_olink_meta_data,
                                           ckbtools_raw_olink_protein_data)

somascan_meta_data <- prepare_somascan_meta_data(ckbtools_raw_somascan_meta_data)

## protein data
olink_protein_data <- prepare_olink_data(ckbtools_raw_olink_protein_data,
                                         olink_meta_data)

somascan_protein_data <- prepare_somascan_data(ckbtools_raw_somascan_protein_data,
                                               ckbtools_raw_somascan_sample_data,
                                               somascan_meta_data,
                                               transform = "log")

## proteomics analysis data
olink_analysis_data <- prepare_proteomics_analysis_dataset(
  subcohort_participant_data,
  olink_protein_data,
  olink_meta_data,
  plate_correction = "none",
  scaling = "scale"
)

somascan_analysis_data <- prepare_proteomics_analysis_dataset(
  subcohort_participant_data,
  somascan_protein_data,
  somascan_meta_data,
  plate_correction = "none",
  scaling = "scale"
)

```


# One model

The `multiple_models()` function can be used to fit a single or multiple linear regression models. The code below will return a the results of a linear regression of protein ACAA1 (Olink) on BMI (kg/m^2^). The model is adjusted for age (and its square), sex, region, fasting time (and its square), outdoor temperature (and its square) and plate.

```{r, message=FALSE}
basic_adjustment_terms <- c("age_at_study_date",
                            "I(age_at_study_date^2)",
                            "sex",
                            "region",
                            "hours_since_last_ate",
                            "I(hours_since_last_ate^2)",
                            "region_mean_temp",
                            "I(region_mean_temp^2)",
                            "plateid")

result <- multiple_models(
  olink_analysis_data,
  model = "linear",
  exposures = "bmi_calc",
  outcomes = "ACAA1",
  adjust_for = list(basic = basic_adjustment_terms)
)

result[, c("outcome", "term", "estimate", "std.error", "p.value", "n")]
```

By default, the function returns one row for a continuous exposure or one row per group for a categorical exposure. Binary or categorical exposures stored as numeric are treated as continuous. These should be changed to factors to treat as categorical.

The code below returns the results of a linear regression of protein CRBB2 (SomaScan) on BMI (grouped) adjusted for age, sex and region.

```{r, message=FALSE}
result <- multiple_models(
  somascan_analysis_data,
  model = "linear",
  exposures = "bmi_grp",
  outcomes = "CRBB2",
  adjust_for = list(basic = basic_adjustment_terms)
)

result[, c("outcome", "term", "estimate", "std.error", "p.value", "n", "n_group")]
```


# Analysis of multiple participant charateristics

Multiple models with different exposures can be specified by setting `exposures` as a character vector.

Use the `remove_adjust_for` argument to remove terms from the adjustment for a specific exposure (or outcome). Here, the age terms are removed from adjustment when the exposure is `age_at_study_date`.

```{r, message=FALSE}
results <- multiple_models(
  olink_analysis_data,
  model = "linear",
  exposures = c("age_at_study_date",
                "sbp_mean",
                "has_diabetes",
                "current_regular_smoker"),
  outcomes = "ACAA1",
  adjust_for = list(basic = basic_adjustment_terms),
  remove_adjust_for = list(age_at_study_date = c("age_at_study_date",
                                                 "I(age_at_study_date^2)"))
)

results[, c("exposure", "term", "estimate", "std.error", "p.value", "n", "n_group")]

```

Note: `has_diabetes` is numeric in the data so is treated as a continuous variable and reported in a single row. Create a factor to treat it is a categorical exposure.  

```{r, message=FALSE}
olink_analysis_data$has_diabetes <- factor(olink_analysis_data$has_diabetes,
                                           levels = c(0, 1),
                                           labels = c("No", "Yes"))

results <- multiple_models(
  olink_analysis_data,
  model = "linear",
  exposures = c("age_at_study_date",
                "sbp_mean",
                "has_diabetes",
                "current_regular_smoker"),
  outcomes = "ACAA1",
  adjust_for = list(basic = basic_adjustment_terms),
  remove_adjust_for = list(age_at_study_date = c("age_at_study_date",
                                                 "I(age_at_study_date^2)"))
)

results[, c("exposure", "term", "estimate", "std.error", "p.value", "n", "n_group")]
```


# Analysis of multiple proteins

Multiple models with different outcomes can be specified by setting `outcomes` as a character vector.

```{r, message=FALSE}
somascan_results <- multiple_models(
  somascan_analysis_data,
  model = "linear",
  exposures = c("sbp_mean", "has_diabetes"),
  outcomes = somascan_meta_data$name,
  adjust_for = list(basic = basic_adjustment_terms)
)
```

# Full analyses

Multiple levels of adjustment are specified using a named list. 

For comparison of nested models the adjustment variables of the simpler model should be included as adjustment factors in the more complex model.   

```{r, message=FALSE}

basic_lifestyle_adjustment_terms <- c(basic_adjustment_terms,
                                      "smoking_3_groups",
                                      "alcohol_3_groups")

olink_results <- multiple_models(
  olink_analysis_data,
  model = "linear",
  exposures = c("sbp_mean", "has_diabetes"),
  outcomes = olink_meta_data$name,
  adjust_for = list(basic = basic_adjustment_terms,
                    basic_lifestyle = basic_lifestyle_adjustment_terms)
)
```


# Check for errors and warnings

After fitting multiple models, always check for warnings and errors. If there are any, decide how these will be handled when summarising results or for further analyses.

```{r}
olink_results %>%
  dplyr::filter(!is.na(error) | !is.na(warnings)) %>%
  dplyr::distinct(outcome, exposure, adjustment, error, warnings)

somascan_results %>%
  dplyr::filter(!is.na(error) | !is.na(warnings)) %>%
  dplyr::distinct(outcome, exposure, adjustment, error, warnings)
```


# Select one result per protein

To summarise the somascan results as in the referenced papers, one aptamer per Uniprot ID was selected choosing the result with the smallest p-value and no error or warning (or just the smallest p-value if all have error or warning).

```{r}
somascan_results <- somascan_results %>%
  dplyr::filter(!is.na(estimate))

somascan_results_per_protein <- somascan_results %>%
  dplyr::inner_join(somascan_meta_data, by = c("outcome" = "name")) %>%
  dplyr::group_by(adjustment, exposure, uniprot) %>%
  dplyr::filter(p.value == min(p.value)) %>%
  dplyr::distinct() %>%
  dplyr::ungroup()
```

In the case of Olink there are a number of assays that are associated with more than one panel.  These are identified in the table below and  

To summarise the Olink results as in the referenced papers, where more than one panel was associated with one assay a single assay was selected across all 4 panels choosing the result with the smallest p-value and no error or warning (or just the smallest p-value if all have error or warning)

```{r}
olink_results_per_protein <- olink_results %>%
  dplyr::filter(!is.na(estimate)) %>%
  dplyr::inner_join(olink_meta_data, by = c("outcome" = "name")) %>%
  dplyr::group_by(adjustment, exposure, uniprot) %>%
  dplyr::filter(p.value == min(p.value)) %>%
  dplyr::distinct() %>%
  dplyr::ungroup()
```
