---
title: "Associations of proteins with incident disease"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Associations of proteins with incident disease}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  class.output = "outp",
  class.message = "outp",
  class.warning = "outp",
  class.error = "outp",
  comment = "#"
)
library(ckbtools)

# set number of Olink plates in synthetic data to prevent warning about number
# of plates
options(olink_expected_plates = 26)
```


This article demonstrates how to use this package to estimate associations between proteins and risk of incidence disease using Cox proportional hazards models. The methods used are from the paper ["Proteome-wide assessment of ~9500 plasma proteins with major diseases and multi-morbidity in a Chinese population"](https://doi.org/10.21203/rs.3.rs-5356315/v1).


# Import data

First, all relevant participant and protein data must be imported into R. For example, use `readRDS()` for RDS format data or `read.csv()` for csv data.
```{r, eval = FALSE}
baseline       <- readRDS("baseline.rds")
ascertainments <- readRDS("ascertainments.rds")
endpoints      <- readRDS("endpoints.rds")

raw_olink_meta_data <- readRDS("olink_explore_meta.rds")
raw_olink_protein_data <- readRDS("data_baseline_olink_explore.rds")

raw_somascan_meta_data <- readRDS("somalogic_meta.rds")
raw_somascan_protein_data <- readRDS("data_baseline_somalogic.rds")
raw_somascan_sample_data  <- readRDS("data_baseline_somalogic_samples.rds")
```

For demonstration, the synthetic data sets available in this package can be used:

1. ckbtools_participant_data
2. ckbtools_ascertainments
3. ckbtools_raw_olink_meta_data
4. ckbtools_raw_olink_protein_data
5. ckbtools_raw_somascan_meta_data
6. ckbtools_raw_somascan_protein_data
7. ckbtools_raw_somascan_sample_data


# Prepare participant data

## Baseline and endpoint data

A data set containing relevant participant baseline and endpoint data must be prepared manually. For these analyses, the data must contain columns for disease endpoints (status and date) and all participant characteristics used for exclusions or as adjustment variables. See [CKB data](ckb_data.html) for an example.


## Select subcohort participants

These analyses of incident disease will use only participants in the subcohort of the nested case-subcohort proteomics study.
```{r}
subcohort_participant_data <- ckbtools_participant_data[ckbtools_participant_data$proteomics_subcohort == 1, ]
subcohort_participant_data$subcohort <- 1
```

## Participant exclusions

Where exclusions are needed for particular analyses, create columns that identify addition participants to be excluded. These will be used later when fitting the survival models. For example, create columns that identify participants to be excluded in analyses of cancer (if they have prior cancer diagnosis at baseline) or analyses of respiratory diseases (if they have prior TB, asthma or COPD at baseline).

```{r}
subcohort_participant_data$exclude_for_cancer <- subcohort_participant_data$cancer_diag == "Yes"
subcohort_participant_data$exclude_for_respiratory <- subcohort_participant_data$tb_diag == "Yes" |
  subcohort_participant_data$asthma_diag == "Yes" |
  subcohort_participant_data$has_copd == "Yes"
```


# Prepare protein data

Prepare the protein data as described in [Proteomics](proteomics.html#preparing-proteomics-data).

For this analysis, Olink values are centred using the plat medians (plate correction "median-subcohort") and SomaScan values are log<sub>2</sub> transformed. Protein values are then scaled to have standard deviations of one.

```{r}
## set CKB data release
options(ckb.data.release = "19.02")

## meta data
olink_meta_data <- prepare_olink_meta_data(ckbtools_raw_olink_meta_data,
                                           ckbtools_raw_olink_protein_data,
                                           names = "id")

somascan_meta_data <- prepare_somascan_meta_data(ckbtools_raw_somascan_meta_data,
                                                 names = "id")

## protein data
olink_protein_data <- prepare_olink_data(ckbtools_raw_olink_protein_data,
                                         olink_meta_data)

somascan_protein_data <- prepare_somascan_data(ckbtools_raw_somascan_protein_data,
                                               ckbtools_raw_somascan_sample_data,
                                               somascan_meta_data,
                                               transform = "log2")

## proteomics analysis data
olink_analysis_data <- prepare_proteomics_analysis_dataset(
  subcohort_participant_data,
  olink_protein_data,
  olink_meta_data,
  plate_correction = "median-subcohort",
  scaling = "scale"
)

somascan_analysis_data <- prepare_proteomics_analysis_dataset(
  subcohort_participant_data,
  somascan_protein_data,
  somascan_meta_data,
  plate_correction = "none",
  scaling = "scale"
)
```


# Analysis of one incident disease and protein

The `cox_model()` function is used to fit one Cox proportional hazards model. The code below will return the fitted proportional hazard models object for an analysis of protein OID13950 with all-cause mortality, with time-on-study as the time scale. The model has baseline hazard stratified by sex and region, and is adjusted for age at baseline and its square, hours since last ate and its square, temperature and its square, and education.

```{r}
cox_model(olink_analysis_data,
          exposure = "OID13950",
          outcome = "all_cause_mortality",
          adjust_for = c("strata(sex)",
                         "strata(region)",
                         "poly(age_at_study_date, 2)",
                         "poly(hours_since_last_ate, 2)",
                         "poly(region_mean_temp, 2)",
                         "education_4_groups"),
          exclude = NULL,
          return = "model")
```

*Note that education_4_groups is formatted as a factor in the data, so it is correctly used as a categorical variable in the model. If you want to include a categorical variable that is coded numerically then make sure to use `"as.factor(variable)"`.*

The fitted model object can be used for further analyses, such as inspection of model residuals, fitted values and fit statistics.


# Analysis of many incident diseases and proteins

The `multiple_models()` function can be used to run multiple regression models.

*It is highly recommended to run some individual analyses as described in the previous section to check that the models and output are as intended, before running analyses for many proteins.*

The code below fits models for all Olink and SomaScan proteins, for two outcomes (`all_cause_mortality` and `respiratory_diseases`) and two sets of adjustments.

For all `respiratory_diseases` models, participants with `exclude_for_respiratory == TRUE` are excluded.

```{r, message=TRUE}
minimally_adjusted <- c("strata(sex)",
                        "strata(region)",
                        "poly(age_at_study_date, 2)",
                        "poly(hours_since_last_ate, 2)",
                        "poly(region_mean_temp, 2)",
                        "education_4_groups")

fully_adjusted <- c("strata(sex)",
                    "strata(region)",
                    "poly(age_at_study_date, 2)",
                    "poly(hours_since_last_ate, 2)",
                    "poly(region_mean_temp, 2)",
                    "education_4_groups",
                    "smoking_3_groups",
                    "alcohol_3_groups",
                    "met",
                    "sbp_mean",
                    "has_diabetes",
                    "bmi_calc")

olink_results <- multiple_models(
  olink_analysis_data,
  model = "cox",
  exposures = olink_meta_data$name,
  outcomes = c("all_cause_mortality",
               "respiratory_diseases"),
  exclude_by_outcome = list(respiratory_diseases = "exclude_for_respiratory"),
  adjust_for = list(minimal = minimally_adjusted,
                    full = fully_adjusted)
)

somascan_results <- multiple_models(
  somascan_analysis_data,
  model = "cox",
  exposures = somascan_meta_data$name,
  outcomes = c("all_cause_mortality",
               "respiratory_diseases"),
  exclude_by_outcome = list(respiratory_diseases = "exclude_for_respiratory"),
  adjust_for = list(minimal = minimally_adjusted,
                    full = fully_adjusted)
)
```

*Note that education, smoking and alcohol are formatted as factors in the data, so they are correctly used as categorical variables in the model. If you want to include a categorical variable that is coded numerically then make sure to use `"as.factor(variable)"`.*

When your analysis includes a very large number of models you can split your analyses into separate uses of the function (e.g. one for each disease endpoint) so that each can be run and saved to a results file before moving onto the next. This `multiple_model()` function or the individual model functions can also be used if you wish to implement another approach to iteration yourself.


# Summarise results

## Check for errors and warnings

After fitting multiple models, always check for warnings and errors. If there are any, decide how these will be handled when summarising results or for further analyses.

```{r}
olink_results %>%
  dplyr::filter(!is.na(error) | !is.na(warnings)) %>%
  dplyr::distinct(outcome, exposure, adjustment, error, warnings)

somascan_results %>%
  dplyr::filter(!is.na(error) | !is.na(warnings)) %>%
  dplyr::distinct(outcome, exposure, adjustment, error, warnings)
```

## Select one result per protein

Some Olink assays are repeated across panels. To summarise the results as in the referenced paper, select the assay by panel preference Cardiometabolic > Inflammation > Neurology > Oncology, and prefer models with no error or warning.

```{r}
olink_summary_results <- olink_results %>%
  dplyr::filter(exposure %in% olink_meta_data$name) %>%
  dplyr::inner_join(olink_meta_data, by = c("exposure" = "name")) %>%
  dplyr::group_by(outcome, adjustment, uniprot) %>%
  dplyr::arrange(!is.na(error) | !is.na(warnings), panel) %>%
  dplyr::filter(dplyr::row_number() == 1) %>%
  dplyr::ungroup()
```

Some SomaScan aptamers target the same protein. To summarise the results as in the referenced paper, select one aptamer per Uniprot ID choosing the result with the smallest p-value and no error or warning (or just the smallest p-value if all have error or warning)

```{r}
somascan_summary_results <- somascan_results %>%
  dplyr::inner_join(somascan_meta_data, by = c("exposure" = "name")) %>%
  dplyr::group_by(outcome, adjustment, uniprot) %>%
  dplyr::arrange(!is.na(error) | !is.na(warnings), p.value) %>%
  dplyr::filter(dplyr::row_number() == 1) %>%
  dplyr::ungroup()
```


## Multiple testing

In the referenced paper, a single multiple testing correction is applied across Olink and SomaScan protein measures for each disease and adjustment model. Where a result has an error or warning, the nominal p-value is first set to one, as the result is not significant but should still be counted as a significance test. In the referenced paper, the false discovery rate correction of [Benjamini & Hochberg (1995)](https://doi.org/10.1111/j.2517-6161.1995.tb02031.x) is used.

```{r}
## combine results from Olink and SomaScan analyses
all_results <- dplyr::bind_rows(olink_summary_results,
                                somascan_summary_results)  %>%
  ## handle errors and warnings
  dplyr::mutate(p.value = dplyr::if_else(!is.na(error) | !is.na(warnings), 1, p.value)) %>%
  ## multiple testing correction per outcome and adjustment
  dplyr::group_by(outcome, adjustment) %>%
  dplyr::mutate(p.value.fdr = p.adjust(p.value, method = "fdr")) %>%
  dplyr::ungroup()
```


# Present results

The number of proteins significantly associated (FDR-corrected p < 0.05) with each disease for each adjustment model can then be summarised.

```{r}
all_results %>%
  dplyr::group_by(outcome, adjustment) %>%
  dplyr::summarise(n = dplyr::n(),
                   error_or_warning = sum(!is.na(error) | !is.na(warnings)),
                   significant = sum(p.value.fdr < 0.05),
                   significant_inverse = sum(p.value.fdr < 0.05 & estimate < 0),
                   .groups = "drop")
```
